Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-10-01 output/breed_paper.Rnw
Revisions from Bruce, Dave, Chris, and Jim

\title{Rice yield improvements through plant breeding are offset by
  inherent yield declines over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David MacKill}
\author[ucd]{Jim E. Hill}
\author[res]{Kent McKenzie}
\author[ucce2]{Michelle Leinfelder-Miles}
\author[ucce]{Luis A. Espino}
\author[ucce]{Randall Mutters}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

@@

\address[res]{Rice Experiment Station, California Cooperative Rice
Research Foundation, Biggs, CA 95917, USA}

@@

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
better understand what improvements have been made in the genetic
yield potential obtained through plant breeding. Specifically, the
hypothesis was tested that the yield of a rice variety declines over
time after the initial selection was tested. This would explain how
side-by-side comparisons show increased yield while no substantive
increases have been seen in farmers' fields. Side-by-side comparisons
under yield erosion will show ``false'' yield increases due to the
yield of the older cultivar decreasing, not due to the increased
genetic yield potential of the newer cultivar. An apparent increase in
yield in side-by-side comparisons between old and new rice varieties
was observed when the yield erosion was not considered. However, this
apparent increase in yield could be explained by the differences in
the time when these cultivars were selected. The impact of yield
erosion was estimated to be 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
credible interval -4.4 to -53.3) after initial selection. Once this
effect was considered, the apparent yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). This underscores the importance of
breeding to maintain yield and the challenges of increasing the
genetic yield potential.
% Add more implications

@@

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 CA state average yields have increased
more slowly (Fig. \ref{fig:state_yield}) despite continuous crop improvement \citep{ccrrf}. For the past 30 years, each newly developed
cultivar has demonstrated a yield advantage over the previous
cultivars in the Statewide Variety Trials (an annual multi-site
variety trial program that compares experimental entries side-by-side
with ``check'' cultivars). However, in a seeming contradiction these same
reports show stagnant yields across release announcements (i.e.,
reports accompanying the office release) from the
various years. For example, at the year of release
(1985), cultivar M-202 was reported to yield an average of 11 Mg
ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the reported average
yield of the newest released cultivar M-209 was 10.8 Mg ha\tsups{-1}
\citep]14\% MC;][]{m209}. In other rice systems, a decrease in a
cultivar's yield performance over time has been observed
\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
erosion \citep{Peng1998}. It has been speculated that this yield erosion is due to
the inability of a selected cultivar's genetics to adapt to changing
biotic and abiotic conditions. To better understand how
plant improvement and yield erosion are contributing to the
improvement of rice yield in farmers' fields, we teseted two competing hypotheses: (1) the
genetic yield advantage increased over the last 30+ years, and (2)
the genetic yield advantage has not increased over the last 30+
because the erosion of yield over time creates an illusion of improved
yields.

@@

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  to \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
  US Dept. of Agriculture National Agricultural Statistics
  Service. Rapid gains in grain yield were experienced from
  approximately 1950 -- 1990, but annual gains since 1990 have been
  slower.}%The line represents the 10 year moving average.}
% Remove 10 year moving average - Bruce does not like. 
\label{fig:state_yield}
\end{figure}

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and conducts the majority of the trials for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially named and released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically planted
in the Statewide Variety Trials for 2 to 4 years prior to their
official release. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected leading to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations representing
\Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release, typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability and the
time elapsed (in years) since release was calculated.

@@

\caption{The yield of released medium grain rice varieties for only
the two years prior to each cultivar's official release. Points
($\bullet$) and bars (|) represent the mean and standard error of the
raw data, respectively. The dashed line and shaded area are the
genetic yield improvement and credible interval as estimated by the model.}
\label{fig:first_yrs}
\end{figure}

@@

Comparing the cultivars' yield during each's initial time in the
experimental trials confirmed that genetic yield improvements have
been consistant over the study period
(Fig. \ref{fig:first_yrs}). As estimated by the full model, there is
no conclusive evidence that the yields of medium grain rice cultivars
selected and released by CCRRF's breeding program have, on average,
increased over the last 30 years (Table \ref{tbl:coef}). The full
model which corrected for both the year of release and yield erosion
was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor. The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

When yield erosion was not corrected for in the comparison
(i.e., omitted from the model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
estimate ignores the confounding effect of yield erosion over time and
performed worse in out-of-sample predictive accuracy, it confirms that
yields in side-by-side comparison would support the conclusion that newer
cultivars out-preform older ones if yield erosion is not considered.

After correcting for the other variables, some cultivars were estimated
to have higher yield compared to the overall average (as estimated by
random intercepts). The marginal estimate for the performance of a
cultivar averaged over the uncertainty in the other parameters in the
model (i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

@@

The average decrease in yield over time once a cultivar is selected
(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
its release. Comparing the two dominate medium grain rice cultivars M-202
(widely planted from the late 1980s until the early 2000s), and M-206
(the most widely planted CCRRFS cultivar at approximately 50\% of
planted area at the time of writing), these two cultivars experienced
similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:coef}). The currently estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
can be attributed solely to the erosion of M-202's yield over time
(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
supporting this estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).

@@

We do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
year\tsups{-1}, both increases and decreases in the genetic yield
potential over the study period are plausible. However, even the
estimated yield improvement that does not account for yield erosion
(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
estimates from this study, there is less than a 1\% probability that
the true value is higher than 45 kg ha\tsups{-1}
year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
has likely been maintaining but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures.

We estimate that yield erosion has decreased the yield of medium grain
rice cultivars released by the CCRRF over the last 30+ years by an
average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
any gains made by the crop improvement program. Biotic and abiotic forces have
previously been speculated to be responsible for yield erosion, since
the cultivar's genetics remain stable from the time of selection but
the conditions the cultivar may experience over time are continuously
changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
is consistent with previous reports of the erosion of genetic gains of
cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
erosion of yield over time is a phenomenon that widely affects rice
breeding programs. This observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit \citep{espe16a}. Without maintenance a breeding
program and continuously selecting cultivars well adapted to current conditions,
CA rice production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

@@

California state average yields have increased slowly
even as little to no gains have been obtained in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, among other management factors. may all be contributing to
increases in the state average yield. However, these gains are far less
than those experienced during the preceeding era, and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) may not be widely adapted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields \citep{espe2017}, and the RES
experiences warmer temperatures than many other rice growing areas in
CA \citep{espe2017}. These cultivars may perform well at the RES but
not widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. However, planning to meet future
demand necessitates increases in grain yield \citep{Godfray2010,
Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
without expansion of land under cultivation will rely on new genetics
with increased yield potential. The evidence here suggests that to
accomplish true increases over time, the improvement target needs take
into account decreases in the current ``standard'' cultivars' yield
over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
continue to be useful, comparisons across years to each cultivar's
benchmark yield at the time of release should provide information on
the absolute improvements being made in yield potential over time. New
technologies may need to be adapted to further increase yield
potential, including hybrid rice.

@@

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                       "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

@@

dotchart(id_est[,1],
         labels = gsub("M", "M-", row.names(id_est)),
         pch = 16,
         xlim = c(min(id_est[,2]),max(id_est[,3])),
         xlab = expression("Estimated difference from average ( Mg ha"^-1~")"))
segments(x0 = id_est[,2], x1 = id_est[,3], y0 = 1:nrow(id_est))
abline(v = 0, lty = 2)
@
\caption{The marginal estimates for the difference between cultivars,
  averaged over the uncertainty in the other parameters.}
\label{fig:ids_comp}
\end{figure}

>>>>

\title{Rice yield improvements through plant breeding are offset by
  inherent yield declines over time}
\author[ucd]{Matthew B. Espe\corref{corresauthor}}
\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
\author[ucd]{David Mackill}
\author[ucd]{Jim E. Hill}
\author[ucce2]{Michelle Leinfelder-Miles}
\author[ucce]{Luis A. Espino}
\author[ucce]{Randall Mutters}
\author[ucd]{Chris van Kessel}
\author[ucd]{Bruce A. Linquist}

@@

@@

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
better understand what improvements have been made in the genetic
yield potential obtained through plant breeding. Specifically, the
hypothesis was tested that the yield of a rice cultivar declines over
time after its introduction. This would explain how
side-by-side comparisons show increased yield while no substantive
increases have been seen in farmers' fields. If there is yield
erosion, side-by-side comparisons will show ``false'' yield increases due to the
yield of the older cultivar decreasing, not due to the increased
genetic yield potential of the newer cultivar. An apparent increase in
yield in side-by-side comparisons between old and new rice varieties
was observed when the yield erosion was not considered. However, this
apparent increase in yield could be explained by the differences in
the time when these cultivars were selected. Yield
eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
credible interval -4.4 to -53.3) after initial selection. Once this
effect was considered, the yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). Thus, without continuous plant improvement and
deployment of new cultivars, yield will not only remain stagnant but
will decline over time. To increasing the genetic yield potential of
rice in the face of this phenomenon, higher yield targets are needed
to counter-act the decrease over time in the current cultivars.

@@

California (CA) rice production represents one such production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly (Fig. \ref{fig:state_yield}), in part due to
improvements in rice genetics (i.e., the ``Green Revolution'') through
the efforts of the California Cooperative Rice Research Foundation
(CCRRF; a collaboration between the University of California, the
USDA-Agricultural Research Service, and farmer-funded
research). However, since 1990 the rate of increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop improvement
\citep{ccrrf}. For the past 30 years, each newly developed cultivar
has demonstrated a yield advantage over the previous cultivars in the
Statewide Variety Trials (an annual multi-site variety trial program
that compares experimental entries side-by-side with ``check''
cultivars). However, in a seeming contradiction these same reports
show stagnant yields across release announcements (i.e., reports
accompanying the official public release) from the previous years. For
example, when cultivar M-202 was released in 1985, it was reported to
yield an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep]14\%
MC;][]{m209}. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed \citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic conditions. To
better understand how plant improvement and yield erosion are
contributing to the improvement of rice yield in farmers' fields, we
teseted the hypotheses: either (1) the genetic yield advantage
increased over the last 30+ years, or (2) the genetic yield advantage
has not increased over the last 30+ because the erosion of yield over
time creates an illusion of improved yields.

@@

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
Agriculture National Agricultural Statistics Service, 2017). Rapid
gains in grain yield were experienced from approximately 1950 -- 1990,
but annual gains since 1990 have been slower.}
\label{fig:state_yield}
\end{figure}

@@

The CCRRF Rice Experiment Station (RES) is located north-west of
Biggs, CA (39.4648, -121.7342), and has been the central location for
CCRRF's efforts to develop improved rice varieties adapted to CA since
1969. The climate at the RES is Mediterranean, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
breeding program, the RES is researcher-managed (as opposed to
farmer-managed trials) and plants the majority of experimental plots for the
Statewide Variety Trials (approximately 3 to 4 times the number of
experimental plots compared to any other location). To avoid potential
complications due to site, climate, and management differences at the
other 6 to 8 farmer-managed trials per year we focused our
investigation on the RES.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. These cultivars are typcically
evaluated in the Statewide Variety Trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
calculated.

@@

\caption{The experimental yield of medium grain rice varieties for the
  two years prior to each cultivar's public release. Points
($\bullet$) and bars (|) represent the mean and standard error of the
raw data, respectively. The dashed line and shaded area are the
genetic yield improvement and credible interval as estimated by the model.}
\label{fig:first_yrs}
\end{figure}

@@

Comparing the cultivars' yield during each's initial time (average 3 years) in the
experimental trials confirmed that genetic yield improvements have
been consistant over the 30 year study period
(Fig. \ref{fig:first_yrs}). As estimated by the model including both
yield erosion and yield improvement over time, there is
no conclusive evidence that the yields of medium grain rice cultivars
selected and released by CCRRF's breeding program have, on average,
increased over the last 30 years (Table \ref{tbl:coef}). The full
model which corrected for both the year of release and yield erosion
was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor (See
supplemental material for further details). The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

When yield erosion was not corrected for in the analysis (i.e.,
omitted from the model), there is an apparent small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
estimate is flawed (i.e., it ignores the confounding effect of yield
erosion over time and performed worse in out-of-sample predictive
accuracy), it confirms previous analyses of side-by-side comparisons
which concluded that newer cultivars out-preform older ones.

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated by random
intercepts). The marginal estimate for the performance of a cultivar
averaged over the uncertainty in the other parameters in the model
(i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

@@

The average decrease in yield over time once a cultivar is selected
(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
its release. The two most dominate medium grain rice cultivars, M-202
(widely planted from the late 1980s until the early 2000s) and M-206
(currently the most widely planted), experienced
similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:coef}). The currently estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
can be attributed solely to the erosion of M-202's yield over time
(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
supporting this estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).

@@

We do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
year\tsups{-1}, both increases and decreases in the genetic yield
potential over the study period are plausible. However, even the
estimated yield improvement that does not account for yield erosion
(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
estimates from this study, there is less than a 1\% probability that
the true value is higher than 45 kg ha\tsups{-1}
year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
has likely maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures.

We estimate that yield erosion has decreased the yield of medium grain
rice cultivars released by the CCRRF over the last 30+ years by an
average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
any gains made by the crop improvement program. Biotic and abiotic forces have
previously been speculated to be responsible for yield erosion, since
the cultivar's genetics remain stable from the time of selection but
the conditions the cultivar may experience over time are continuously
changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
is consistent with previous reports of the erosion of genetic gains of
cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
erosion of yield over time is a phenomenon that widely affects rice
breeding programs. This observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit \citep{espe16a}. Without maintenance breeding
(i.e., continuously selecting cultivars well adapted to current
conditions), CA rice production systems are predicted to have lost an
estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
yield over the study period.

@@

California state average yields have increased slowly
despite there being little to no gains in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, and other factors may all be contributing to
increases in the state average yield. However, these gains are far less
than those experienced during the preceeding era, and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adapted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields \citep{espe2017}, and the RES
experiences warmer temperatures than many other rice growing areas in
CA \citep{espe2017}. These cultivars may perform well at the RES but
not widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality \citep{mckenzie1993} and
milling yield \citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate varieties grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table 2). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration varieties would presumably require less irrigation
water).

Planning to meet future
demand necessitates increases in grain yield \citep{Godfray2010,
Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
without expansion of land under cultivation will rely on new genetics
with increased yield potential. The evidence here suggests that to
accomplish true increases over time, the improvement target needs take
into account decreases in the current ``standard'' cultivars' yield
over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
continue to be useful, comparisons across years to each cultivar's
benchmark yield at the time of release should provide information on
the absolute improvements being made in yield potential over time. New
technologies may need to be adapted to further increase yield
potential, including hybrid rice.

@@

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

@@

dotchart(id_est[,1],
         labels = gsub("M", "M-", row.names(id_est)),
         pch = 16,
         xlim = c(min(id_est[,2]),max(id_est[,3])),
         xlab = expression("Estimated difference from average ( Mg ha"^-1~")"))
segments(x0 = id_est[,2], x1 = id_est[,3], y0 = 1:nrow(id_est))
abline(v = 0, lty = 2)
@
\caption{The marginal estimates for the difference between cultivars,
  averaged over the uncertainty in the fixed effect parameters (i.e.,
  yield erosion and annual yield gains).}
\label{fig:ids_comp}
\end{figure}
